Two subtle problems with over-representation analysis

### 2024-06-13 manuscript/manuscript.Rmd
ms

TODO.

@@

### Methods

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected methods for nine different
gene set libraries.
The mean number of significant results was calculated for each of the gene set libraries
before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
increase was observed for HPO (229%).
The similarity between results, as quantified with the Jaccard similarity was highest for
Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).

@@

We posited that the foreground gene list length might exacerbate the problem, so we performed
parallel analysis with foreground lists of length between 125 and 2,000 genes (**Figure 2B**).
A clear trend was observed for GO, Reactome, Cellmarkers and TFT GTRD, with Jaccard similarity
being significantly reduced when foreground gene lists were shorter than 500 genes.
Interestingly, the drop in Jaccard similarity was more severe for GO and Reactome sets as
compared to Cellmarkers and TFT GTRD sets, and may be associated with the size of the gene
sets in the library.
Median gene set size for these are GO:18, Reactome:11, TFT GTRD:287 and Cellmarkers:115.

@@

Simulated differential expression profiles with apriori changes to predetermined pathway genes were
analysed in parallel by clusterProfiler's `enricher` function, which suffers from the two
aforementioned problems and `fora` which does not.
In parallel, `fgsea` was used to compare the performance of these two ORA methods against a
prototypical FCS method.
ClusterProfiler scored very high precision when added variance was less than 0.4, but showed a
progressive worsening in precision as more noise was added to the dataset, dropping to 0.61 with added
variance of 0.6.
In comparison, the precision of fora was relatively stable around 0.97 until added variance reached 0.5.
The recall of ORA methods dropped suddenly when added variance was 0.2 or larger.
fgsea showed stable precision of 0.94 when added variance was 0.4 or lower, and dropped slightly to
0.91 with added variance of 0.6.
fgsea recall was better than either ORA tool across all levels of added variance.
This result shows that these two subtle problems result in poorer precision for clusterProfiler,
which becomes more severe with higher levels of noise.
Overall, fgsea had similar precision to fora (mean=0.94), but the higher recall of fgsea meant it had
a higher overall accuracy measured by F1 score (0.83) as compared to fora and clusterProfiler
(0.61 and 0.64 respectively).

@@

Here we have described the effect of two subtle problems with common implementations of ORA.
This work was inspired by a doctoral student new to bioinformatics who was puzzled by confusing results
obtained from clusterProfiler when using a custom set of pathways.
Our hunch that clusterProfiler was behaving in an undesired way was confirmed by the discovery of
several posts to online bioinformatics forums dating to ~2020 where confused users were wondering why
the number of background genes changed so much.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated genes from
the background `options(enrichment_force_universe = TRUE)`, however this feature isn't yet described
in the official documentation.

The second problem is potentially more serious in that it could lead large numbers false positives
under specific conditions; when the foreground list is short and the median pathway size is relatively
small.
These false positives arise due to an underestimate of the actual number of tests conducted.
ClusterProfiler is not the only tool to suffer from these problems, they are widespread (**Table 1**).

@@

Although the purpose of this work was to describe two subtle problems with ORA, it has some major
drawbacks which limit its utility.

* The method is rarely reported comprehensively, especially crucial details about background gene list
and FDR correction.

* Results vary dramatically depending on the number of genes selected in the foreground.

* Overall ORA is less accurate as compared to FCS

These bigger problems suggest FCS should replace ORA in most cases.
Where it makes sense to use ORA would be when FCS is not applicable, for example when investigating
the over-represented categories in gene expression modules defined by packages like WGCNA (Weighted
Gene Correlation Network Analysis)[@Langfelder2008-hu].

>>>>

Over-representation analysis (ORA) is used widely to asses for the enrichment of functional
categories in a gene list compared to a background list.
ORA is therefore a critical method in the interpretation of 'omics data, relating gene lists
to biological functions and themes.
Although ORA is hugely popular, we and others have noticed two potentially undesired
behaviours of some ORA tools.
The first one we call the "background problem", because it involves the software eliminating
large numbers of genes from the background list if they are not annotated as belonging to any
category.
The second one we call the "false discovery rate problem", because some tools underestimate
the true number of parallel tests conducted.
Here we demonstrate the impact of these issues on several real RNA-seq datasets and use
simulated RNA-seq data to quantify the impact of these problems.
We show that the severity of these problems depends on the gene set library, the number of
genes in the list, and the degree of noise in the dataset.
These problem can be mitigated by changing packages/websites for ORA or by changing to another
approach such as functional class scoring.

@@

## Methods

@@

A workaround to eliminate the background problem was developed and we used it to compare the
results of ORA of seven datasets with the original and corrected methods for nine different
gene set libraries.
The mean number of significant results was calculated for each of the gene set libraries
before and after correcting the problem (**Figure 1A**).
Each of the gene set libraries showed an uplift when correcting the background problem.
The smallest increase was observed in Cellmarkers gene library (11.7%), while the largest
increase was observed for HPO (229%).
The similarity between results, as quantified with the Jaccard similarity was highest for
Cellmarkers and GTRD at 0.88 and 0.76 respectively, and the lowest similarity was recorded for
HPO and KEGG with 0.20 and 0.48 respectively.
Observed Jaccard scores correlated with the number of genes annotated to one or more functional
categories in the gene set library (**Figure 1B**).
This result suggests that ORA with smaller gene set libraries like HPO and KEGG are impacted
more severely.

@@

We posited that the foreground gene list length might exacerbate the problem, so we performed
parallel analysis with foreground lists of length between 125 and 2,000 genes (**Figure 2B**).
A clear trend was observed for GO, Reactome, Cellmarkers and TFT GTRD, with Jaccard similarity
being significantly reduced when foreground gene lists were shorter than 500 genes.
Interestingly, the drop in Jaccard similarity was more severe for GO and Reactome sets as
compared to Cellmarkers and TFT GTRD sets, and may be associated with the size of the gene
sets in the library.
Median gene set size for these are GO:18, Reactome:11, TFT GTRD:287 and Cellmarkers:115.
These results demonstrate that the FDR problem is more severe when dealing with shorter
foreground lists, and this is exacerbated when functional categories in the gene set library
are also small.

@@

Simulated differential expression profiles with *a priori* changes to predetermined pathway
genes were analysed in parallel by clusterProfiler's `enricher` function, which suffers from
the two aforementioned problems and `fora` which does not.
In parallel, `fgsea` was used to compare the performance of these two ORA methods against a
prototypical FCS method.
ClusterProfiler scored very high precision when added variance was less than 0.2, but showed a
progressive worsening in precision as more noise was added to the dataset, dropping to 0.69
with added variance of 0.5.
In comparison, the precision of fora was relatively stable around 0.97 until added variance
reached 0.5, where it dropped to 0.88.
The recall of ORA methods dropped suddenly when added variance was 0.2 or greater.
fgsea showed stable precision of 0.93 or greater when added variance was 0.5 or lower, which
dropped slightly to 0.91 with added variance of 0.6.
fgsea recall was better than either ORA tool across all levels of added variance.
This result shows that these two subtle problems result in poorer precision for
clusterProfiler, which becomes more severe with higher levels of noise.
Overall, fgsea had similar mean precision to fora (0.94), but the higher recall of fgsea
meant it had a higher overall accuracy measured by F1 score (0.83) as compared to fora and
clusterProfiler (0.61 and 0.64 respectively).

@@

The background problem came to our attention when a doctoral student in our department
was puzzled by results obtained from clusterProfiler when using a custom set of pathways.
This exact issue was the subject of several posts to online bioinformatics forums dating to
~2020 where confused users were wondering why the number of background genes changed so much.
The results shown here show that the background problem effects smaller gene set libraries
like KEGG and HPC more severely as compared to Cellmarkers and GO libraries which describe
a larger proportion of all genes.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated
genes from the background `options(enrichment_force_universe = TRUE)`, however this feature
isn't yet described in the official documentation.

The second problem is potentially more serious in that it could lead to increasing false
positives under specific conditions; when the foreground list is short and the median pathway
size is relatively small.
These false positives arise due to an underestimate of the actual number of tests conducted.
In extreme cases, correction of this problem can lead to around half of the significant
results changing.

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 1**), but there are also several tools free of these problems.

@@

Although the purpose of this work was to describe two subtle problems with ORA, this work
highlights some major drawbacks.
Firstly, ORA was less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Secondly ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.
These issues together with previously mentioned pitfalls such as lack of background
correction and methodological misreporting [@Timmons2015-ki;@Wijesooriya2022-fw], suggest
that ORA should be discouraged in most situations, and replaced by other methods such as FCS.
Where it makes sense to use ORA would be when FCS is not applicable, for example when
investigating the over-represented categories in gene expression modules defined by packages
like WGCNA (Weighted Gene Correlation Network Analysis)[@Langfelder2008-hu], or when
investigating over-represented functional categories among genes with variants associated
with a disease.
